---
title: "03-AL-Baseline-Model"
author: "Geri"
date: "2026-05-14"
output: html_document
---

## Setup

```{r}
install.packages("ordinal")
library(ordinal)
library(MASS)
library(splines)
library(ggplot2)
library(tidyr)
library(dplyr)
library(broom)
```

## Load and prepare data

```{r}
df <- read.csv("swan_analysis_ready_v3.csv")

df$sleep_quality <- ordered(df$sleep_quality)
df$meno_stage    <- factor(df$meno_stage, ordered = TRUE)
df$SWANID        <- factor(df$SWANID)
df$RACE          <- factor(df$RACE)
df$presc_med     <- factor(df$presc_med, levels = c(0, 1), labels = c("No", "Yes"))
df$VMS_binary    <- as.integer(df$VMS_binary)
df$INCOME        <- factor(df$INCOME, levels = c(1,2,3,4),
                           labels = c("<20k","20-49k","50-99k","100k+"))
```

---

## STEP 1: Fit the primary model (clmm with random intercept)
# nAGQ = 10 uses a more accurate integration method than the default Laplace.

```{r}
model_clmm2_baseline_controlled_sp <- clmm(
  sleep_quality ~ AL_score + meno_stage + AGE_baseline + RACE + presc_med +
    INCOME + VMS_binary + ns(CESD_score, df = 3) + (1 | SWANID),
  data      = df,
  link      = "logit",
  Hess      = TRUE,
  nAGQ      = 10,
  control   = clmm.control(maxIter = 100, gradTol = 1e-4),
  na.action = na.omit
)


summary(model_clmm2_baseline_controlled_sp)
```
```{r}
summary(model_clmm2_baseline_controlled_sp)

```
```{r}
exp(coef(model_clmm2_baseline_controlled_sp)) #odds ratios
exp(confint(model_clmm2_baseline_controlled_sp)) #confidence intervals
```



```{r}
ci <- confint(model_clmm2_baseline_controlled_sp)

OR_table <- cbind(
  OR = exp(coef(model_clmm2_baseline_controlled_sp)),
  exp(ci)
)

OR_table
```
```{r}
pacman::p_load(ggeffects)

pred <- ggpredict(
  model_clmm2_baseline_controlled_sp,
  terms = "AL_score"
)

plot(pred)
```


# Quick look at odds ratios from the primary model
# exp(coef) converts log-odds to odds ratios
# A value > 1 means higher odds of worse sleep; < 1 means lower odds

```{r}
exp(model_clmm2_controlled$beta)
```

---

## STEP 2: Check the proportional odds assumption

# The proportional odds assumption says the effect of each predictor is the
# same across all three thresholds of sleep quality. We test this using
# nominal_test(), which requires a clm (no random effect) version of the model.

```{r}
model_clm2 <- clm(
  sleep_quality ~ AL_score + meno_stage + AGE_baseline + RACE + presc_med +
    INCOME + VMS_binary + CESD_score,
  data      = df,
  link      = "logit",
  na.action = na.omit
)

nominal_test(model_clm2)
```

# If any predictors are flagged (p < .05), they are to be moved into the nominal 
# argument below. This fits a partial proportional odds model where those predictors 

```{r}
model_clm2_partial <- clm(
  sleep_quality ~ AL_score + meno_stage + AGE_baseline + RACE + presc_med +
    INCOME + VMS_binary + CESD_score,
  nominal   = ~ meno_stage + RACE + CESD_score + presc_med + 
                INCOME,
  data      = df,
  link      = "logit",
  na.action = na.omit
)

nominal_test(model_clm2_partial)
```

# Compare AL_score coefficient between constrained and partial model.
# If they are close, the violations in the covariates were not distorting
# the main estimate — the finding is robust.

```{r}
cat("AL_score in constrained clm:    ", coef(model_clm2)["AL_score"], "\n")
cat("AL_score in partial clm:        ", coef(model_clm2_partial)["AL_score"], "\n")
cat("AL_score in primary clmm:       ", model_clmm2_baseline_controlled_sp$beta["AL_score"], "\n")
```

---

## STEP 3: Predicted probability plot for AL score

# This is the main results plot. For each AL score value (0-8), it shows the
# predicted probability of each sleep quality category, holding all other
# predictors at typical/reference values.

```{r}
newdat_al <- expand.grid(
  AL_score     = 0:8,
  meno_stage   = levels(df$meno_stage)[2],           # early peri as reference
  AGE_baseline = mean(df$AGE_baseline, na.rm = TRUE),
  RACE         = levels(df$RACE)[1],
  presc_med    = "No",
  INCOME       = "20-49k",
  VMS_binary   = 0,
  CESD_score   = mean(df$CESD_score, na.rm = TRUE)
)

# Predict() does not work directly on clmm, so I use the clm version
# The pattern will be the same; the clmm just additionally accounts for
# between-person variance via the random intercept
pred_al <- predict(model_clm2, newdata = newdat_al, type = "prob")
pred_df <- cbind(newdat_al, as.data.frame(pred_al$fit))

pred_long <- pivot_longer(pred_df,
                          cols      = c("1","2","3","4"),
                          names_to  = "sleep_quality",
                          values_to = "probability")

ggplot(pred_long, aes(x = AL_score, y = probability,
                      colour = sleep_quality, group = sleep_quality)) +
  geom_line(linewidth = 1) +
  geom_point() +
  scale_x_continuous(breaks = 0:8) +
  labs(x      = "Allostatic Load Score",
       y      = "Predicted Probability",
       colour = "Sleep Quality",
       title = "Predicted probabilities of sleep quality categories by baseline allostatic load") +
  theme_minimal()
```


```{r}
poor_sleep_df <- pred_long %>%
  mutate(sleep_quality = as.numeric(sleep_quality)) %>%
  group_by(AL_score) %>%
  summarise(
    poor_sleep_probability = sum(probability[sleep_quality %in% c(3, 4)]),
    .groups = "drop"
  )

theme_thesis <- theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11, colour = "grey35"),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    legend.position = "bottom",
    axis.title = element_text(face = "bold")
  )

ggplot(poor_sleep_df, aes(x = AL_score, y = poor_sleep_probability)) +
  geom_line(linewidth = 1.2, colour = "#2C3E50") +
  geom_point(size = 2.8, colour = "#2C3E50") +
  scale_x_continuous(breaks = 0:8) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    x = "Baseline allostatic load score",
    y = "Predicted probability",
    title = "Predicted probability of poorer sleep quality",
    subtitle = "Poorer sleep defined as sleep quality categories 3–4"
  ) +
  theme_thesis
```
```{r}
df %>%
  filter(!is.na(AL_score), !is.na(sleep_quality)) %>%
  count(AL_score, sleep_quality) %>%
  group_by(AL_score) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = AL_score, y = prop, fill = sleep_quality)) +
  geom_col(width = 0.8, colour = "white", linewidth = 0.2) +
  scale_x_continuous(breaks = 0:8) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_viridis_d(
    option = "C",
    direction = -1,
    name = "Sleep quality category"
  ) +
  labs(
    x = "Baseline allostatic load score",
    y = "Observed proportion",
    title = "Observed sleep quality distribution",
    subtitle = "Distribution of sleep quality categories by baseline allostatic load score"
  ) +
  theme_thesis
```


## STEP 4: Check linearity of CESD score

# Tests whether the relationship between CESD and sleep quality is linear
# or whether it curves. Uses a natural spline as the flexible alternative.
# anova() compares model fit — a significant result means non-linearity.

```{r}
model_linear <- polr(
  sleep_quality ~ CESD_score + AL_score + meno_stage +
    AGE_baseline + RACE + presc_med + INCOME + VMS_binary,
  data      = df,
  Hess      = TRUE,
  na.action = na.omit
)

model_spline <- polr(
  sleep_quality ~ ns(CESD_score, df = 3) + AL_score + meno_stage +
    AGE_baseline + RACE + presc_med + INCOME + VMS_binary,
  data      = df,
  Hess      = TRUE,
  na.action = na.omit
)

anova(model_linear, model_spline)
```

# Plot predicted probabilities across the CESD range using the spline model
# to visualise the shape of the relationship

```{r}
newdat_cesd <- data.frame(
  CESD_score   = seq(min(df$CESD_score, na.rm = TRUE),
                     max(df$CESD_score, na.rm = TRUE),
                     length.out = 100),
  AL_score     = mean(df$AL_score, na.rm = TRUE),
  meno_stage   = levels(df$meno_stage)[2],
  AGE_baseline = mean(df$AGE_baseline, na.rm = TRUE),
  RACE         = levels(df$RACE)[1],
  presc_med    = levels(df$presc_med)[1],
  INCOME       = levels(df$INCOME)[2],
  VMS_binary   = 0
)

pred_cesd <- predict(model_spline, newdata = newdat_cesd, type = "probs")

cols <- c("black", "red", "green", "blue")
matplot(newdat_cesd$CESD_score, pred_cesd,
        type = "l", lty = 1, lwd = 2, col = cols,
        xlab = "CES-D score", ylab = "Predicted probability",
        main = "Predicted sleep quality probabilities across CES-D score")
legend("right", legend = colnames(pred_cesd),
       col = cols, lty = 1, lwd = 2, bty = "n")
```
```{r}
cat("AL_score linear: ", coef(model_linear)["AL_score"], "\n")
cat("AL_score spline: ", coef(model_spline)["AL_score"], "\n")
```


##Checking the linnearity of the continuous AL variables using already-established ns(CESD_score) as base
```{r}

# log_DHAS — only for the DHEAS model
anova(
  polr(sleep_quality ~ AL_score + meno_stage + AGE_baseline + RACE + 
         presc_med + INCOME + VMS_binary + ns(CESD_score, df=3), 
       data=df, Hess=TRUE, na.action=na.omit),
  polr(sleep_quality ~ ns(AL_score, df=3) + meno_stage + AGE_baseline + RACE + 
         presc_med + INCOME + VMS_binary + ns(CESD_score, df=3), 
       data=df, Hess=TRUE, na.action=na.omit)
)
```

```{r}
# Then AGE_baseline linearity check
model_age_linear <- polr(
  sleep_quality ~ AL_score + meno_stage + AGE_baseline + RACE + presc_med +
    INCOME + VMS_binary + ns(CESD_score, df = 3),
  data = df, Hess = TRUE, na.action = na.omit
)

model_age_spline <- polr(
  sleep_quality ~ AL_score + meno_stage + ns(AGE_baseline, df = 3) + RACE + presc_med +
    INCOME + VMS_binary + ns(CESD_score, df = 3),
  data = df, Hess = TRUE, na.action = na.omit
)


anova(model_age_linear, model_age_spline)

# Also check coefficient stability
cat("AL_score linear AGE:  ", coef(model_age_linear)["AL_score"], "\n")
cat("AL_score spline AGE:  ", coef(model_age_spline)["AL_score"], "\n")
```

```{r}
modelsRDS <- list(
  baseline_controlled_sp = model_clmm2_baseline_controlled_sp,
  DHEAS_controlled_sp = model_clmm2_DHEAS_controlled_sp,
  long_controlled_sp = model_clmm2_long_controlled_sp,
  reduced_controlled_sp = model_clmm2_reduced_controlled_sp
)

saveRDS(modelsRDS, "clmm_models.rds")
```


---

## ARCHIVE: Earlier exploratory models (not reported)

# Simple model kept only to track observations lost to missingness
# compared to the full model above

```{r}
model_clmm <- clmm(
  sleep_quality ~ AL_score + meno_stage + AGE_baseline + (1 | SWANID),
  data      = df,
  link      = "logit",
  Hess      = TRUE,
  na.action = na.omit
)

summary(model_clmm)
```

# Version with time-varying AGE instead of AGE_baseline
# VIF confirmed no collinearity issue but AGE_baseline kept for consistency
# with primary model specification

```{r}
model_clmm3_controlled <- clmm(
  sleep_quality ~ AL_score + meno_stage + AGE + RACE + presc_med +
    INCOME + VMS_binary + CESD_score + (1 | SWANID),
  data      = df,
  link      = "logit",
  Hess      = TRUE,
  nAGQ      = 10,
  control   = clmm.control(maxIter = 100, gradTol = 1e-4),
  na.action = na.omit
)

summary(model_clmm3_controlled)
```
